---
title: "P1805_HW3_kp2809"
author: "Kimberly Lopez"
date: "2024-10-07"
output: github_document
---
Loading Data 
```{r include=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)
library(p8105.datasets)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

```

```{r}
data("ny_noaa")
nrow(ny_noaa)
```

The ny_noaa data set has 7 columns and 2.5 million row entries which detail the id, date of observed data, and precipitation (tenths of mm), snow (mm), snow depth (mm), temperature max(tenths of degrees C), and temperature min (tenths of degrees C). Id being Weather station ID name.There is some missing data in columns for precipitation, snwd, tempertaure max and min. 


Do some data cleaning. Create separate variables for year, month, and day. Ensure observations for temperature, precipitation, and snowfall are given in reasonable units. For snowfall, what are the most commonly observed values? Why?

- date has year, month, and day. Need to pivot wider and save into 3 new columns 

```{r}
ny_noaa = 
  ny_noaa |>
  separate(date, into = c("year", "month", "day"), convert = TRUE) |> 
  mutate(
    tmax = as.numeric(tmax),
    tmin = as.numeric(tmin))


```

```{r}
ny_noaa |> 
  select(snow)|>
  drop_na()|>
  pull(snow)|> 
  range()
```
The range of snow fall is -13 to 10,160. 

```{r eval=FALSE, include=FALSE}
ny_noaa|>
  ggplot(aes(x= id,y = snow)) + 
  geom_point()
```

*Temperature* 
```{r eval=FALSE, include=FALSE}
ny_noaa |> 
  select(tmax)|>
  drop_na()|> 
  pull(tmax)|>
  range()

```
The range of temperature is -1-99 which could be in F rather than Celcius. 
```{r eval=FALSE, include=FALSE}
ny_noaa|>
  ggplot(aes(x= id,y = prcp)) + 
  geom_point()
```

The range for precipitation is 22860 which leads me to think they are not in the same units. 
There appears to be 4 outliers in reported snowfall which might be calculated in different units. 
Make a two-panel plot showing the average max temperature in January and in July in each station across years. Is there any observable / interpretable structure? Any outliers?


```{r}
ny_noaa |> 
  group_by(id, year, month) |> 
  filter(month %in% c(1, 7)) |> 
  summarize(mean_tmax = mean(tmax, na.rm = TRUE, color = id)) |> 
  ggplot(aes(x = year, y = mean_tmax, group = id)) + 
  geom_point() + 
  geom_path() +
  facet_grid(~month) +
  labs(title = "Average monthly temperature for each station across years for January and July")
```


Make a two-panel plot showing (i) tmax vs tmin for the full dataset (note that a scatterplot may not be the best option); and (ii) make a plot showing the distribution of snowfall values greater than 0 and less than 100 separately by year.

```{r}
hex = 
  ny_noaa |>
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_hex()
hex
```

```{r}
ridge = 
  ny_noaa |>
  filter(snow < 100, snow > 0) |>
  ggplot(aes(x = snow, y = as.factor(year))) + 
  geom_density_ridges()

hex + ridge
```



# Problem 2 

Load, tidy, merge, and otherwise organize the data sets. Your final dataset should include all originally observed variables; 

- exclude participants less than 21 years of age, and those with missing demographic data
- Might need to pivot longer saving all min1 times to a column of minute_day and saving the values to a column of phys_activity for each of those times
- encode data with reasonable variable classes (i.e. not numeric, and using factors with the ordering of tables and plots in mind).

```{r}
nhanes_accel = read_csv("data/nhanes_accel.csv",na = c("NA","","."))
nhanes_covar = read_csv("data/nhanes_covar.csv", na = c("NA","","."), skip =4)

nhanes_df = 
  inner_join(nhanes_accel, nhanes_covar, by= "SEQN")|>
  janitor::clean_names()|>
  relocate(seqn, sex, age, bmi, education)|>
  filter(age >=21)|>
  drop_na(bmi,education)

nhanes_df
```
Produce a reader-friendly table for the number of men and women in each education category, and create a visualization of the age distributions for men and women in each education category. Comment on these items.

- 1= male, 2= female
- 2 rows men or women 
- 3 columns sex, education level 1, 2, 3 

```{r}
nhanes_df |>
  select(sex, education) |>
  table() |>
  knitr::kable(col.names = c("Less than HS", "HS Equivalent", "More than HS"))

```

This table shows us there are 27 males with less than a HS education, 35 with HS equivalent education, and 56 with more than a HS level education. It also shows 28 females with less than HS, 23 with HS equivalent level of education, and 59 with more than a HS level education in this sample. 


Aggregated across minutes to create a total activity variable for each participant. 

```{r}
nhanes_df=
  nhanes_df |>
  mutate(
    tot_activ = rowSums(select(nhanes_df, min1:min1440), na.rm = TRUE))
nhanes_df
```

Below is a plot of these total activities (y-axis) against age (x-axis) and compare men to women and have separate panels for each education level.

```{r}

female = 
  nhanes_df|>
  filter(sex==2)
male= nhanes_df|>
  filter(sex==1)

nhanes_df |>
  ggplot(aes(x= age, y = tot_activ, color = sex))+ 
  geom_point (alpha = .3)  + 
  facet_grid(. ~ education)+
  geom_smooth(data = female, se=FALSE)+ 
  geom_smooth(data=male, se=FALSE)+ 
  labs(
    title = "Total activity during a 24-hour interval",
    x = "Age",
    y = "Total Acitivity",
    color = "sex",
    caption = "Total Acitivty Data Across Education Levels and Different Sexes"
  )
  
```

This plot shows that across sexes there are similar trends in total activity by age. Younger age groups had higher total activity compared to older age individuals regardless of sex in each individual education level. This shows a decrease in total physical activity as adults get older. However, compared across education level, youth with less than a high school education for both male and females had higher total activity in a 24-hour time period compared to younger adults with a higher than high school education. There is greater variation in total activity among those with an education higher than high school compared to the other groups. 


Below is a three-panel plot that shows the 24-hour activity time courses for each education level and uses color to indicate sex. Describe in words any patterns or conclusions you can make based on this graph; including smooth trends may help identify differences.

Pivoting time columns to long to have an easier column to plot on x axis for time.
```{r}
nhanes_df_min = 
  inner_join(nhanes_accel, nhanes_covar, by= "SEQN")|>
  janitor::clean_names()|>
  relocate(seqn, sex, age, bmi, education)|>
  filter(age >=21)|>
  drop_na(bmi,education)|>
  pivot_longer(
    cols= min1:min1440,
    names_to = "minute_in_day", 
    names_prefix = "min",
    values_to= "phys_activ"
  )|>
  mutate(
    minute_in_day = as.numeric(minute_in_day)
  )
  
```

```{r}
nhanes_df_min |>
  ggplot(aes(x= minute_in_day, y = phys_activ, color = sex))+ 
  geom_point (alpha = .05)  +
  geom_smooth(data = filter(nhanes_df_min, sex==2), se=FALSE)+ 
  geom_smooth(data=filter(nhanes_df_min, sex==1), se=FALSE)+
  facet_grid(. ~ education)+ 
  viridis::scale_color_viridis() + 
  theme_minimal() + 
  theme(legend.position = "bottom")+ 
  labs(
    title = "Physical Acitivity Overtime During a 24-hour interval",
    x = "Minutes in a day",
    y = "Physical Acitivity",
    color = "Sex",
    caption = "Accelerometer data")
```
This plot shows physical activity did not have much variation in trends across different sex. There were higher peaks in physical activity among those who have an education level more than high school. Overtime, all 3 education levels have dips in physical activity 



# Problem 3 

Load City Bike data and combined all datasets with year and month tags.

```{r}
citi_2020_jan= read_csv("data/citibike/Jan 2020 Citi.csv.zip", na = c("NA","","."))|>
  janitor:: clean_names()|>
  mutate(
    year= 2020,
    month= "jan"
  )
citi_2020_july = read_csv("data/citibike/July 2020 Citi.csv.zip", na = c("NA","","."))|>
  janitor:: clean_names()|>
  mutate(
    year= 2020,
    month= "july"
  )

citi_2024_jan= read_csv("data/citibike/Jan 2024 Citi.csv.zip",, na = c("NA","","."))|>
  janitor:: clean_names()|>
  mutate(
    year= 2024,
    month= "jan"
  )
citi_2024_july= read_csv("data/citibike/July 2024 Citi.csv.zip", na = c("NA","","."))|>
  janitor:: clean_names()|>
  mutate(
    year= 2024,
    month= "july"
  )

citi_df= 
  bind_rows(citi_2020_jan, citi_2020_july,citi_2024_jan,citi_2024_july)|> 
  janitor:: clean_names()|>
  relocate(year,month)

```


1.) Produce a reader-friendly table showing the total number of rides in each combination of year and month separating casual riders and Citi Bike members. Comment on these results.

```{r}

citi_df |>
  group_by(year, month) |>
  summarize(
    member = sum(member_casual == "member"),
    casual_rider = sum(member_casual == "casual")
  )|>
  knitr::kable()

```

The number of casual riders from January 2020 has increased overtime to be similar in proportion to member riders by July 2024. Both casual riders and members have increased since 2020. There is normally more riders in general during July than Janurary of both years likley due to weather conditions during that time of year favoring bike riding. 


2.) Make a table showing the 5 most popular starting stations for July 2024; include the number of rides originating from these stations.

```{r}

popularity =
  citi_df |>
  filter(month == "july", year == 2024) |>
  select(start_station_name) |>
  table()

popularity = 
  popularity |>
  as.data.frame()|>
  arrange(desc(Freq))|>
  head(n=5)

popularity
```

3.) Make a plot to investigate the effects of day of the week, month, and year on median ride duration. This plot can include one or more panels, but should facilitate comparison across all variables of interest.

```{r}
month= citi_df|>
  ggplot(aes(x=as.factor(month), y = duration))+ 
  geom_boxplot()

day= 
  citi_df |> 
  mutate(
    weekdays = factor(weekdays, levels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")))|>
  ggplot(aes(x= as.factor(weekdays), y = duration ))+
  geom_boxplot()+
  theme(axis.text.x = element_text(angle = 45, size = 10, margin = margin(t = 10)))

year= 
  citi_df|>
  ggplot(aes(x= as.factor(year), y = duration)) + 
  geom_boxplot()

day+month + year
```

We can see that the data is skewed by outliers that pull the range for duration close to 250 minutes. We can see a slight dip in the middle of the week for median ride duration, and can see slightly higher ride duration on the weekends. Comparing by month, median ride duration is higher in July than January and there is a higher IQR of variation in duration in July. 2020 showed a higher median duration compared to 2024. In 2024, we also saw smaller IQR in duration compared to 2020. Across all time variables, median fluctuates; increasing in warmer months, decreasing in recent year, and decreasing in the middle of the weekdays.


4.) There were relatively few electric Citi Bikes in 2020, but many more are available now. For data in 2024, make a figure that shows the impact of month, membership status, and bike type on the distribution of ride duration. Comment on your results.

```{r}
month_duration=
  citi_df|>
  filter(year==2024)|>
  ggplot( aes(x = duration, y =month)) + 
  geom_density_ridges()

member_duration= 
  citi_df|>
  filter(year==2024)|>
  ggplot( aes(x = duration, y =member_casual)) + 
  geom_density_ridges()

bike_duration = 
  citi_df|>
  filter(year==2024)|>
  ggplot( aes(x = duration, y =rideable_type)) + 
  geom_density_ridges()

month_duration/ member_duration / bike_duration
  
```

The density plot helps to visualize the distribution of duration, a continuous variable. We can see from the plots, again, that there are multiple outliers. If we only focus on the left where the data is most dense, we notice that the peak for membership is much higher than causal riders. This highlights the positive impact of membership status on the amount of riders, but the peak slightly occurs before the causal riders. This means that casual riders have slightly longer duration in rides. In terms of months, January appears to have a peak which occurs slightly shorter in duration than in July. Those who used the classic bike had a density peak which occurred lower on the x - axis, duration, compared to electric bikes. 



